On the performance of the MAC model

### 2018-10-31 2018_nips_vigil/mac.tex
Mac and S-Mac rewrite

The MAC network~\cite{hudson2018compositional} is a recurrent model that performs
sequential reasoning, where each step involves analyzing a part of the question following 
which attention on the image is changed suitably.
At the end, a suitable answer is given based on this composite reasoning.
Both the question and image are transformed suitably prior to the reasoning process. 
For the question, a sequence of vectors obtained via word-embeddings is fed
to a bidirectional LSTM to produce a sequence of hidden state pairs. Each pair 
corresponds to a unique word in the question and denotes its 
\emph{contextual embedding}. The final states in each direction 
represent the \emph{global context} of that question.
Similarly each image is initially processed using a pre-trained network (using ResNet~\cite{he2016resnet}) to obtain a feature map. 
This is used as an input to a trainable neural network that consists of 
two CNN layers followed by a single linear layer\footnote{%
This minor reformulation saves computation time instead of applying it
in \emph{every} reasoning step.}.
This results in a 3D tensor for each image in the data set collectively called the
\emph{knowledge base}~\cite{hudson2018compositional}.

@@

The core of the network is a sequence of recurrent MAC cells. A MAC cell (Figure 1) is performing one reasoning step. At every step the network is driving the attention on one part of the question and the corresponding part of the image. This operation is made possible by the combination of 3 units working together, a control unit, a read and a write units. The control unit is updating the control state and drives the attention over the question. Given this control state, the read unit then retrieves information from the memory and drives the attention over the image.
Finally, the write unit retrieves this visual information and updates the memory state.
The last part of the network is the output unit. The output unit is simply a classifier that process the concatenation of the question representation and the final memory to produce a final answer.

>>>>

The MAC network~\cite{hudson2018compositional} is a recurrent model that performs
sequential reasoning, where each step involves analyzing a part of the question following 
which attention on the image is changed suitably.
At the end, a suitable answer is given based on this composite reasoning.
Both the question and image are transformed suitably prior to the reasoning process. 
For the question, a sequence of vectors obtained via word-embeddings is fed
to a bidirectional LSTM to produce a sequence of hidden state pairs. Each pair 
corresponds to a unique word in the question and denotes its 
\emph{contextual embedding}. The final states in each direction 
represent the \emph{global context} of that question.
Similarly each image is initially processed using a pre-trained network (using ResNet~\cite{he2016resnet}) to obtain a feature map. 
This is given as an input to a trainable neural network that consists of 
two CNN layers, resulting in a 3D tensor for each image in the data set. By viewing this
tensor a collection of 1D vectors in the last dimension, we call this
the \emph{knowledge base} of that image.

@@

The core of the network is a sequence of recurrent MAC cells, each performing one reasoning step (see Figure~\ref{fig:mac_cell}). At every step the network is driving the attention on one part of the question and the corresponding part of the image. This operation is made possible by the combination of 3 units working together: a control unit, a read unit and a write unit. The control unit is updating the control state and drives the attention over the question. Given this control state, the read unit then retrieves information from the memory and drives the attention over the image.
Finally, the write unit retrieves this visual information and updates the memory state.
The last part of the network is the output unit,  a simple classifier that process the concatenation of the global context  and the final memory state to produce a final answer.

### 2018-10-31 2018_nips_vigil/smac.tex
tiny changes
Mac and S-Mac rewrite

\subsection{Simplified MAC network}
Our proposed modification to the MAC network is based on two heuristic
simplifications of the equations governing a MAC cell. 
First, we observed that when looking at the equations of different units
together as a whole, we notice that there are several affine transformations
that are applied in sequence (with no activation in-between).
Therefore they can actually be combined to form a single affine transformation.
The second observation uses the following heuristic argument:
suppose there is a single affine transformation which preserves the number
of dimensions. Then one can assume that the transformation is invertible
so as to avoid a loss in the information. Applying this principle to 
equations in the read and write unit results in eliminating an affine
operation of the knowledge base in the recurrent network. Instead we compute
this operation on the knowledge base right after we apply the 2 
trainable CNN layers (i.e. prior to the reasoning steps). This results
in a significant savings in training time, as our experiments demonstrate
below, without incurring too much drop in accuracy.

@@

\noindent\textit{Notation:} The index/position $i$ below denotes the reasoning step. 
Unless mentioned otherwise, the weights/biases are invariant w.r.t $i$, 
i.e., shared across all the reasoning steps.

\noindent\textbf{Control unit:} 
For both models, in the control unit, the question $q$ is first transformed in each step of 
%the reasoning using a \emph{position-aware}~\cite{hudson2018compositional} 
linear layer: $q_i = U_i^{[d \times 2d]} q + b_i^{[d]}$.

@@

\noindent\textbf{Read and write units:}
For both models below, $\kb_{hw}$ denotes the 
\emph{knowledge base}~\cite{hudson2018compositional}
consisting of 3D tensors with dimension $H \times W \times d$ for each image
in the data set. The difference in computing $\kb_{hw}$ is that the simplified MAC uses
an additional linear layer on top of the 2 CNN layers present in the original MAC network.
This circumvents the need for performing additional linear transformations on the 
knowledge base, as is done by the original MAC network in \emph{each} reasoning step.

\begin{multicols}{2}
	\noindent
	\begin{align*}
	&I_{ihw} = (W_{m}^{[d \times d]} \mem_{i-1} + b_{m}^{[d]}) \\
	           & \qquad \quad \odot (W_{k}^{[d \times d]} \kb_{hw} + b_{k}^{[d]}) \tag{r1} \\
	&I'_{ihw} =  W_{I'}^{[d \times 2d]} [I_{ihw},\kb_{hw}]  + b_{I'}^{[d]}  \tag{r2} \\
	&ra_{ihw} = W_{ra}^{[1 \times d]} (\cc_i \odot I'_{ihw}) + b_{ra}^{[1]} \tag{r3.1}\\
	&rv_{ihw} = \textrm{softmax}(ra_{ihw}) \tag{r3.2}\\
	&\rr_i = \sum_s rv_{ihw} \, \kb_{hw}  \tag{r3.3}\\
	&\mem_i = W_{rm}^{[d \times d]} [\rr_i, \mem_{i-1}]  + b_{rm}^{[d]} \tag{w1}	
	\end{align*}
	\columnbreak
	{\color{Plum}
	\begin{align*}
	&I_{ihw} = \mem_{i-1} \odot \kb_{hw} \tag{r1} \\ \\
	&I'_{ihw} = W_{I'}^{[d \times d]} I_{ihw} + b_{I'}^{[d]} + \kb_{hw} \tag{r2} \\
	&ra_{ihw} = W_{ra}^{[1 \times d]} (\cc_i \odot I'_{ihw})  \tag{r3.1}\\
	&rv_{ihw} = \textrm{softmax}(ra_{ihw}) \tag{r3.2}\\
	&\rr_i = \sum_s rv_{ihw} \, \kb_{hw}  \tag{r3.3}\\
	&\mem_i = W_{rm}^{[d \times d]} \rr_i + b_{rm}^{[d]} \tag{w1}
	\end{align*}}
\end{multicols}

\begin{table}
	\centering
	\begin{tabular}{ccccCcCc}
		\toprule
	     Model        & Read Unit               & Write Unit &  Control Unit         \\
		\midrule
		MAC   &  787,969 &  524,800        &    6,822,913    \\
		%\midrule
	    simplified MAC & 263,168  & 262,656       &    6,560,768 \\
	    \midrule
	     Reduction [\%]  & -67\%  &   -50\%       &      -4\%  \\
		\bottomrule
	\end{tabular}
	\caption{Numbers of trainable parameters in the units of MAC and simplified MAC cells}
	\label{tab:parameters}
\end{table}

>>>>

\subsection{Simplified MAC network}
Our proposed modification to the MAC network is based on two heuristic
simplifications of the MAC cell. 
First, we observe that when looking at the equations of different units
together as a whole, there are several linear layers
that are applied in sequence (with no activation in-between).
Therefore they can actually be combined to form a single linear layer.
The second observation uses the following heuristic assumption:
suppose there is an affine transformation that preserves the number
of dimensions. Then one can assume that the transformation is invertible
so as to avoid a loss in the information. Applying this principle to 
equations in the MAC cell results in eliminating one of the linear layers applied to
the knowledge base in the cell. With a careful reorganization, we in fact
can compute this operation on the knowledge base right after we apply the 2 
 CNN layers on the feature maps (i.e. prior to the reasoning steps). Thus, we can work purely
with this \emph{projected} knowledge base, that yields
noticeable savings in training time, as our experiments demonstrate
below, without incurring too much drop in accuracy.

@@

\noindent\textit{Notation.}
Let $q$ denote the global context of the question
and let $\cw_s$ denote the contextual embedding of a word $s$ in the question.
For each tensor of dimension $H \times W \times d$ corresponding to an image,
its knowledge base $\{ \kb_{hw}\}$ is the set of $d$-dimensional vectors indexed by 
$h \in \{1,2,\dots,H\}$ and $w \in \{1,2,\dots, W\}$.
The index/position $i$ below denotes the reasoning step. 
Unless mentioned otherwise, the weights/biases are invariant w.r.t $i$, 
i.e., shared across all the reasoning steps.

\noindent\textbf{Control unit:} 
For both models, in the control unit, the question $q$ is first transformed in each step of 
the reasoning using a \emph{position-aware}
linear layer depending on $i$: $q_i = U_i^{[d \times 2d]} q + b_i^{[d]}$.

@@

\noindent\textbf{Read and write units:}
\begin{multicols}{2}
	\noindent
	\begin{align*}
	&I_{ihw} = (W_{m}^{[d \times d]} \mem_{i-1} + b_{m}^{[d]}) \\
	           & \qquad \quad \odot (W_{k}^{[d \times d]} \kb_{hw} + b_{k}^{[d]}) \tag{r1} \\
	&I'_{ihw} =  W_{I'}^{[d \times 2d]} [I_{ihw},\kb_{hw}]  + b_{I'}^{[d]}  \tag{r2} \\
	&ra_{ihw} = W_{ra}^{[1 \times d]} (\cc_i \odot I'_{ihw}) + b_{ra}^{[1]} \tag{r3.1}\\
	&rv_{ihw} = \textrm{softmax}(ra_{ihw}) \tag{r3.2}\\
	&\rr_i = \sum_s rv_{ihw} \, \kb_{hw}  \tag{r3.3}\\
	&\mem_i = W_{rm}^{[d \times d]} [\rr_i, \mem_{i-1}]  + b_{rm}^{[d]} \tag{w1}	
	\end{align*}
	\columnbreak
	{\color{Plum}
	\begin{align*}
	&I_{ihw} = \mem_{i-1} \odot \kb_{hw} \tag{r1} \\ \\
	&I'_{ihw} = W_{I'}^{[d \times d]} I_{ihw} + b_{I'}^{[d]} + \kb_{hw} \tag{r2} \\
	&ra_{ihw} = W_{ra}^{[1 \times d]} (\cc_i \odot I'_{ihw})  \tag{r3.1}\\
	&rv_{ihw} = \textrm{softmax}(ra_{ihw}) \tag{r3.2}\\
	&\rr_i = \sum_s rv_{ihw} \, \kb_{hw}  \tag{r3.3}\\
	&\mem_i = W_{rm}^{[d \times d]} \rr_i + b_{rm}^{[d]} \tag{w1}
	\end{align*}}
\end{multicols}
For the original MAC model shown on the left above, eqn.~(r1) can be slightly optimized for time
by computing the linear layer on the knowledge base \emph{prior} to the reasoning steps.
This induces additional space due to storing this projected knowledge base throughout the
reasoning steps.

\begin{table}
	\centering
	\begin{tabular}{ccccCcCc}
		\toprule
	     Model        & Read Unit               & Write Unit &  Control Unit         \\
		\midrule
		MAC   &  787,969 &  524,800        &    6,822,913    \\
		%\midrule
	    simplified MAC & 263,168  & 262,656       &    6,560,768 \\
	    \midrule
	     Reduction by [\%]  & 67\%  &   50\%       &      4\%  \\
		\bottomrule
	\end{tabular}
	\caption{Numbers of trainable parameters in the units of MAC and simplified MAC cells}
	\label{tab:parameters}
\end{table}
